Analyse gender in music psychology from journal publication data

### 2025-09-27 manuscript/manuscript.qmd
Changed title. Small tweaks in discussion.

---
title             : "Gender Distribution of Authors in Music Psychology"
shorttitle        : "Gender in Music Psychology"
author: 
  - name          : "Tuomas Eerola"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Palace Green, Department of Music, Durham University, DH1 3RL Durham, United Kingdom"
    email         : "tuomas.eerola@durham.ac.uk"
  - name          : "Anna Czepiel"
    affiliation   : "2"
    address       : "Department of Psychology, University of Toronto Mississauga, Canada"
    email         : "a.czepiel@utoronto.ca"
affiliation:
  - id            : "1"
    institution   : "Department of Music, Durham University"
  - id            : "2"
    institution   : "Department of Psychology, University of Toronto Mississauga"

@@

Looking more broadly past the information within papers themselves, other research provides some further explanations for discrepancies in female/male authorships. Despite efforts to give fair credit for author contributions [@ni2021gendered] using, for example, the Contributor Roles Taxonomy (CRediT) system, assigning authorship can be unclear. Women are more likely to experience authorship disagreements, for example, when to decide authorship: women prefer to discuss in earlier stages, while men choose authorship at the final stage [@ni2021gendered]. Another reason for the gender gap could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Additionally, women faculty perform more university, campus, and departmental services than men in general [@guarino2017faculty]. Fortunately, there are strides to improve the current situation, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison]. To improve academic equity issues, it is worthwhile focusing on subdiscipline to provide more 'localised' support to improve the equity in field. This was the aim of the current paper to look within our own field of music psychology. Indeed, the past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

# Discussion

Authorship is a crucial marker of scholarly success in academia, which strongly influences academic career progression, recognition, funding, and advancement opportunities. However, there are several inequities within academic authorship, with women often being underrepresented. As male-female authorship representations differ across different academic disciplines, it is worthwhile diagnosing the situation in the field of music psychology. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender in core music psychology journals, specifically _Musicae Scientiae_, _Psychology of Music_, _Music Perception_, _Journal of New Music Research_, and _Music & Science_. Overall, we found that the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology [34.8% in @eerola2026inprep] and higher than in psychology more broadly, where women account for a third of authors [33.2% in @huang2020; 36.2%in @gonzalez2020women]. This representation is also much higher than in other STEM subjects of math, physics and computer science [15% in @huang2020] and topics such as astronomy and bioscience [17.8% in @son2022scientific]. While our current findings suggests that the field of music psychology is doing better in fostering gender equity than certain other fields, women are still overall underrepresented. In exploring the nuances behind this, we further analysed differences between authorship role, number of citations, geography, and article keywords.

In terms of authorship roles, notable differences emerge. Women are more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This underrepresentation in woman as last authors corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Although last authorship in the social sciences is not necessarily universally defined, it is nonetheless increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution]. Therefore, this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This patterns is also consistent with the idea of a "leaky pipeline", where women are more likely to drop away from academic careers with increasing seniority, especially between doctoral and (consolidated) researcher positions [@gonzalez2020women] and generally have shorter academic careers [@sanchez2024analysis]. On a broader level, this might highlight the need for more mentorship programs for (junior) women to senior positions. Indeed, mentorship programs have been shown support remaining in academia [@gardiner2007show]. We note that within the broader music psychology community, there are some mentorship programs in place, for example, Women in Music Information Retrieval WIMIR (https://wimir.wordpress.com/) and the Society for Music Perception and Cognition has resources for trainees and a Mentorship committe (https://www.musicperception.org/committees) and often holds mentor-mentee meetings at conferences. Although these measures have been in places for some years, it might still take some time for the effect to translate and additional mentorship programs could be welcome.

@@

Turning to potential drivers of these gender differences, geographical and cultural factors appear to play a major role. Our summaries of gender patterns across continents and countries indicate clear cross-national variation, supporting the idea that certain inequity may stem from national cultural factors [@bruck2023bibliometric]. Large Commonwealth nations, namely Australia, Canada, and the United Kingdom, seem to provide a cultural climate more supportive of women in research, as reflected in the relatively high odds ratios of women appearing as authors (in any authorship position) in music psychology. This pattern, however, is not universal across all countries examined [@chan2020]. Prior work shows that women’s research success can be positively correlated with national indicators of gender equity [@chan2020] or are related to countries with intermediate Human Development Index [@sanchez2024analysis; @sugimoto2015relationship]. This discrepancy between author representation could likely to be related to a career length, which strongly correlates with overall productivity. Women exhibit higher dropout rates than men, particularly during the 5–10 years following their first publication [@huang2020], but policies geared towards equity and support for career breaks and caring responsibilities may well contribute towards the observed country-specific differences. Nevertheless, prior analyses demonstrate that even when women constitute a smaller share of researchers within a discipline or country, their contributions are often more impactful than those of their male colleagues [@chan2020]. When examining countries where women have significantly lower odds of authorship compared to men (Japan, Belgium, Austria, Sweden, and Turkey), it is not straightforward to conclude that the lower proportion of female author is related gender equality rankings. The notable exception is Turkey, which is positioned lower according to the Gender Inequality Index [@UNDPHDR2024].

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on (health)care and nurturing [@bruck2023bibliometric; @gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. The difference that female prefer 'caring' topics over male authors fits with the general idea that female researchers value research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. This fits with another potential explanation for the disparity between authorship, namely that women might also lean toward and execute more university/campus/departmental services than men [@guarino2017faculty]. Over time, performing such services to universities are time away from other work on acquiring funding and leading research, which would aggregate over time to time away from publishing research. It is nonetheless important for such services are completed; in some cases it could be there is more encouragement for males to take on these kinds of roles, but also it could be that success in academia could also be more societal defined as not only publications [@kiai2019protect], but as a whole should celebrate achievements that support internal, local and community services in a broader sense.

There are certain limitations of the current research. The gender of the authors are not part of the database - gender was inferred based on author name using _genderize API_ [@wais2006]. Nonetheless, we maintain that the conclusions may still hold. First, both authors checked whether any of the gender attribution of major names needed fixing (see accompanying code). Second, there error (confidence) of the gender attribution was relatively small (only `89` had low (< 0.55) attribution probability, while 89.3% had high (> 0.90) attribution probability), so it is unlikely that the remaining names would significantly change the main conclusions. Another limitation is treating gender as a binary category. While we recognise that gender is a complex and multidimensional social construct, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and because generally the number of non-binary authors is currently relatively low [0.8% in @son2022scientific]. Nonetheless, future research should also move towards using more inclusive, non-binary information. However, the main limitation is that such research indicates the symptoms of certain issues in academia, but not necessarily the causes. While we speculate on certain reasons for the current state of affairs, other kinds of research such as qualitative and longitudinal research would further elucidate the key problems, and where areas for improvements could be made to promite better equity in academia.

<!-- I did not know what these references signaled, so temporarily disabled them: -->

<!-- @ashmos2011gendering : women publish more qualitative studies than men -->

<!-- @sebo2020gender : women were more likely to publish qualitative studies and to publish in primary healthcare journals -->

<!-- @thelwall2019gender : women are likely to use exploratory and qualitative methods rather than quantitative methods. -->

# Conclusions

The analysis of gender distributions in the five most prominent music psychology journals over the last 25 years presented a rich and evolving history of gender distribution in various authorship roles. While overall women hold smaller number of authorship overall (`{r} round(nrow(dplyr::filter(df,Gender=='female'))/nrow(df),3)*100`%), they are significantly more likely to hold first author positions in coauthored publications but significantly less likely likely to hold last author positions than men. For solo publications and being a coauthor, there are no significant gender differences.

@@

While we present analysis of gender of the authorships across the discipline-specific journal in a relatively simple manner here, there would be more elaborate ways into obtaining insights into the gendered career progression in music psychology; for instance, future work with the same dataa could analyse the precarity of career options as judged from the years of authorship attributions [see @lundine2018]. One could also analyse the last author positions in more detail using direct contacts to the authors (interviews and surveys) and try to identify the factors contributing to the barriers experienced by women to take up senior author positions.

Current findings from authorship counts in music psychology present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia in general are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are broadly comparable to psychology as a whole, musicology and its closely related disciplines such as music theory, popular music studies, and ethnomusicology tend to show less balanced gender representation in overall authorship. Gender, however, is only one dimension of equity; racial, ethnic, geographical, institutional, and other inequalities also merit attention in music psychology. Although women remain underrepresented in terms of absolute authorship counts in specialist music psychology journals, even a simple tally of contributions can help open discussion and provide evidence of progress toward greater inclusion and equality. To interpret these patterns more fully, one must examine how music psychology operates through its journals, conferences, training programmes, and curricula, as well as how the discipline presents itself both in countries with a long-standing tradition in the field and in regions where it has yet to become firmly established.

>>>>

---
title             : "The Gendered Nature of Authorship in Music Psychology"
shorttitle        : "Gender in Music Psychology"
author: 
  - name          : "Tuomas Eerola"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Palace Green, Department of Music, Durham University, DH1 3RL Durham, United Kingdom"
    email         : "tuomas.eerola@durham.ac.uk"
  - name          : "Anna Czepiel"
    affiliation   : "2"
    address       : "Department of Psychology, University of Toronto Mississauga, Canada"
    email         : "a.czepiel@utoronto.ca"
affiliation:
  - id            : "1"
    institution   : "Department of Music, Durham University"
  - id            : "2"
    institution   : "Department of Psychology, University of Toronto Mississauga"

@@

To improve academic equity issues, it is worthwhile focusing on subdiscipline to provide more 'localised' support to improve the equity in field. This was the aim of the current paper to look within our own field of music psychology. Indeed, the past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

# Discussion
A crucial marker of scholarly success in academia is authorship, which strongly influences career progression, recognition, funding, and advancement opportunities. However, there are several inequities within academic authorship, where women are often underrepresented. As male-female authorship representations differ across different academic disciplines, the goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender in the field of music psychology. An analysis of papers from five core music psychology journals across 25 years revealed that, overall, women represented 40.2% of authors. This proportion is higher than in musicology [34.8% in @eerola2026inprep], in psychology more broadly [33.2% in @huang2020; 36.2% in @gonzalez2020women], and also much higher than in other STEM subjects of math, physics and computer science [15% in @huang2020] and topics such as astronomy and bioscience [17.8% in @son2022scientific]. Although our current findings suggests that the field of music psychology is doing better in fostering gender equity than some other fields, the distribution of authors by gender remains nonetheless unequal. In exploring the nuances behind this, we further analysed differences between authorship role, number of citations, geography, and article keywords.

In terms of authorship roles, notable authorship differences emerge. Women are more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This is in line with previous research showing that female authorship representation is increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, we found that women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This underrepresentation in woman as last authors is similar to the situation in medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Although last authorship in the social sciences is not necessarily universally defined, it is nonetheless increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution]. Therefore, this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This pattern is also consistent with finding that women are more likely to drop away from academic careers with increasing seniority [@gonzalez2020women; @sanchez2024analysis]. Indeed, this gender gap could come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Fortunately, there are strides to improve the current situation, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison]. In particular, one crucial aspect that has been highlighted to is the need for more mentorship programs for (junior) women to senior positions. Indeed, mentorship programs can encourage mentees to remain in academia [@gardiner2007show]. We note that within the broader music psychology community, there are already mentorship programs in place: for example, Women in Music Information Retrieval WIMIR (https://wimir.wordpress.com/) and the Society for Music Perception and Cognition (SMPC) has resources for trainees and a Mentorship committee (https://www.musicperception.org/committees) and often holds mentor-mentee meetings at conferences. Although these measures have been implemented for some years, it might still take some time for the effect to be reflected in authorship.

@@

In terms of eographical and cultural factors, our summaries of gender patterns across continents and countries indicate clear cross-national variation, supporting the idea that certain inequity may stem from national cultural factors [@bruck2023bibliometric]. Large Commonwealth nations, namely Australia, Canada, and the United Kingdom, seem to provide a cultural climate more supportive of women in research, as reflected in the relatively high odds ratios of women appearing as authors (in any authorship position) in music psychology. This pattern, however, is not universal across all countries examined [@chan2020]. Prior work shows that women’s research success can be positively correlated with national indicators of gender equity [@chan2020] or are related to countries with intermediate Human Development Index [@sanchez2024analysis; @sugimoto2015relationship]. This discrepancy between author representation could likely to be related to a career length, which strongly correlates with overall productivity. Women exhibit higher dropout rates than men, particularly during the 5–10 years following their first publication [@huang2020], but policies geared towards equity and support for career breaks and caring responsibilities may well contribute towards the observed country-specific differences. Nevertheless, prior analyses demonstrate that even when women constitute a smaller share of researchers within a discipline or country, their contributions are often more impactful than those of their male colleagues [@chan2020]. When examining countries where women have significantly lower odds of authorship compared to men (Japan, Belgium, Austria, Sweden, and Turkey), it is not straightforward to conclude that the lower proportion of female author is related gender equality rankings. The notable exception is Turkey, which is positioned lower according to the Gender Inequality Index [@UNDPHDR2024].

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on (health)care and nurturing [@bruck2023bibliometric; @gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. The difference that female prefer 'caring' topics over male authors fits with the general idea that female researchers value research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. This fits with another potential explanation for the disparity between authorship, namely that women faculty lean word and perform more university, campus, and departmental services than men in general [@guarino2017faculty]. Over time, performing such services to universities are time away from other work on acquiring funding and leading research, which would aggregate over time to time away from publishing research. It is nonetheless important for such services are completed; in some cases it could be there is more encouragement for males to take on these kinds of roles, but also it could be that success in academia could also be more societal defined as not only publications [@kiai2019protect], but as a whole should celebrate achievements that support internal, local and community services in a broader sense.

There are certain limitations of the current research. First, the gender of the authors are not part of the database - gender was inferred based on author name using _genderize API_ [@wais2006]. Nonetheless, we maintain that the current conclusions may still hold as both authors checked whether any of the gender attribution of major names needed fixing (see accompanying code). Additionally, gender attribution error was relatively small (only `89` had low (< 0.55) attribution probability, while 89.3% had high (> 0.90) attribution probability), so it is unlikely that the remaining names would significantly change the main conclusions. A second limitation is treating gender as a binary category. While we recognise that gender is a complex and multidimensional social construct, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and because generally the number of non-binary authors is currently relatively low [0.8% in @son2022scientific]. Nonetheless, future research should also move towards using more inclusive, non-binary information. The third - and main - limitation is that such bibliometric research indicates the symptoms of academia issues, but not necessarily the causes. While we speculate on certain reasons for the current state of affairs as discussed, other kinds of research such as qualitative and longitudinal research would further elucidate the key problems, and where areas for improvements could be made to promote better equity in academia.

# Conclusions
The analysis of gender distributions in the five most prominent music psychology journals over the last 25 years presented a rich and evolving history of gender distribution in various authorship roles. While overall women hold smaller number of authorship overall (`{r} round(nrow(dplyr::filter(df,Gender=='female'))/nrow(df),3)*100`%), they are significantly more likely to hold first author positions in coauthored publications but significantly less likely likely to hold last author positions than men. For solo publications and being a coauthor, there were no significant gender differences.

@@

While we present analysis of gender of the authorships across the discipline-specific journal in a relatively simple manner here, there would be more elaborate ways into obtaining insights into the gendered career progression in music psychology; for instance, future work with the same data could analyse the precarity of career options as judged from the years of authorship attributions [see @lundine2018]. One could also analyse the last author positions in more detail using direct contacts to the authors (interviews and surveys) and try to identify the factors contributing to the barriers experienced by women to take up senior author positions.

Current findings from authorship counts in music psychology present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia in general are [@bhopal2021competing; @harris2025gender; @Filandri03082021]. While the results are broadly comparable to psychology as a whole, musicology and its closely related disciplines such as music theory, popular music studies, and ethnomusicology tend to show less balanced gender representation in overall authorship. Gender, however, is only one dimension of equity; racial, ethnic, geographical, institutional, and other inequalities also merit attention in music psychology. Although women remain underrepresented in terms of absolute authorship counts in specialist music psychology journals, even a simple tally of contributions can help open discussion and provide evidence of progress toward greater inclusion and equality. To interpret these patterns more fully, one must examine how music psychology operates through its journals, conferences, training programmes, and curricula, as well as how the discipline presents itself both in countries with a long-standing tradition in the field and in regions where it has yet to become firmly established.
